The genomic impacts of drift and selection for hybrid performance in maize

### 2015-08-27 main.tex
edits to discussion -- jri finalish draft

Reciprocal recurrent selection serves as a model for the method of hybrid maize improvement \citep{duvick2004long} and similar patterns of diversity and population structure can be seen broadly across North American maize germplasm \citep{van2012historical}, suggesting , genetic drift has most likely played a large role in the current genetic structure of modern maize.  	
Several key inbreds in the ‘stiff-stalk’ heterotic group, B73, B37 and B14, were derived from the BSSS population \citep{darrah1986, troyer1999background}. 
B37 and B14 were derived from cycle 0, and B73 was derived from a half-sib recurrent selection program also started with the BSSS population. 
We examined these three inbreds at the pericentromeric regions listed in Table \ref{tab:fix}, and found that in most cases they carry different haplotypes from those that rose to high frequency in the RRS experiment.  
Near isogenic lines that substitute these centromeric haplotypes from key inbreds could be used to determine if the haplotypes at high frequency in cycle 16 of BSSS provide a phenotypic advantage when crossed with BSCB1. \jri{removed paragraph on drift at cycle 0. shortens discussion, felt out of place and we mention both possibilities already in the results}

%Both our study and previous work \citep{labate1997molecular} identified genetic differentiation between cycle 0 of BSSS and the founder inbreds, \rev{likely reflecting the loss of some rare alleles.
%If this differentiation reflects drift during population maintenance before initiation of the RRS, it may have impacted the trajectory of the RRS program and its derived inbreds. 
%However, cycle 0 has also undergone maintenance since the beginning of the RRS experiment, so some} drift may have occurred during this time as well.

Although drift can explain most of the genetic structure genome-wide, phenotypic data provide clear evidence that selection has altered the frequencies of favorable alleles in the BSSS and BSCB1 populations. 
Numerous experiments have shown that the selected populations and the hybrids formed from them \rev{exhibit genetic gain for hybrid yield as well plant architecture and tolerance to high-density planting} \citep{smith1983evaluation, keeratinijakal1993responses, schnicker1993interpopulation, holthaus1995population, brekke2011selection, brekke2011selectionb, edwards2011changes, lauer2012morphological}. 
Since these same trends are observed across wider North American germplasm \citep{tollenaar1989genetic}, identification and characterization of loci conferring phenotypic improvement in BSSS and BSCB1 could reveal some of the genetic mechanisms by which maize yield has steadily improved over the decades. 
We find that heterozygosity falls more than expected across the genome as a whole, and, \rev{though drift imposes limitations on the power to detect selection at individual loci,} genomic regions of extremely low diversity evident at cycle 16 are unlikely to be produced by simulation.  
We further show that an identity-by-descent, haplotype-based approach provides additional power to identify selected regions, as it can distinguish between the fixation of rare and common haplotypes.

\citet{van2012historical} assessed variation in and tested for selection across a wide range of North American maize germplasm using the same SNP array. 
In spite of the large genetic gains made in decades of maize breeding \citep{duvick2005contribution}, \citet{van2012historical} identified  few loci showing  selected loci were identified in that study, the overall effects on genomic diversity and haplotype patterns were relatively small compared with the effects of genetic drift. 
One explanation for the limited overall impact of selection could arise because different haplotypes and loci are selected in different breeding programs. 
In support of this idea, we find that the most likely targets of selection between the BSSS and BSCB1 populations occur at non-overlapping loci. 
This non-overlap between putatively selected loci in complementary populations has also been observed in commercial breeding programs \citep{feng2006temporal}.

The observation that the same targets of selection are not observed in the opposing heterotic populations bears implications for the genetic mechanisms responsible for heterosis and the success of maize hybrids. 
Classic overdominance models of heterosis predict that at a single locus, two distinct alleles confer heterozygote advantage when combined, while the dominance model predicts that heterosis is driven by dominance effects and the complementation of linked alleles in low-recombination regions (dominance or pseudo-overdominance). 
In the case of true overdominance,  selection should lead to decreased heterozygosity at a locus in both populations as complementary haplotypes are fixed in each group \rev{\citep[e.g.][]{guo2014maize}}.
We find \rev{little evidence to support this genetic phenomenon, finding only two 2 cM windows genome-wide in which both populations show significantly reduced heterozygosity.} 
The observed pattern instead favors a dominance model, where fixation of a haplotype in one population simply selects against that same haplotype in the other population. 
\rev{Although strongly deleterious variants were likely purged during the inbreeding process leading to the founder lines, many weakly deleterious alleles can be found segregating at low frequencies among inbreds \citep{Mezmouk2014}.}
Because deleterious alleles will be rare in both populations, most haplotypes in the second population will have a different suite of deleterious variants and will complement the fixed haplotype reasonably well.
\rev{Selection against homozygosity of the fixed haplotype will thus} have little impact on diversity in the second population.

Our data are consistent with such a process having been important for hybrid improvement in the Iowa RSS experiment, and the lack of extreme differentiation seen over time across US Corn Belt germplasm is consistent with a role for such selection in maize hybrid improvement in general \citep{van2012historical}.

@@

>>>>

Although drift can explain most of the genetic structure genome-wide, phenotypic data provide clear evidence that selection has altered the frequencies of favorable alleles in the BSSS and BSCB1 populations. 
Numerous experiments have shown that the selected populations and the hybrids formed from them \rev{exhibit genetic gain for hybrid yield as well plant architecture and tolerance to high-density planting} \citep{smith1983evaluation, keeratinijakal1993responses, schnicker1993interpopulation, holthaus1995population, brekke2011selection, brekke2011selectionb, edwards2011changes, lauer2012morphological}. 
We find that heterozygosity falls more than expected across the genome as a whole, and, \rev{though drift imposes limitations on the power to detect selection at individual loci,} genomic regions of extremely low diversity evident at cycle 16 are unlikely to be produced by simulation.  
We further show that an identity-by-descent, haplotype-based approach provides additional power to identify selected regions, as it can distinguish between the fixation of rare and common haplotypes. 
\rev{These analyses show that most likely targets of selection occur at different loci in the two populations, a result consistent with analyses in commercial breeding programs \citep{feng2006temporal} and which may help explain the lack of selection seen in previous analyses across numerous breeding programs \citep{van2012historical}.}

The observation that different targets of selection are  observed in  opposing heterotic populations bears implications for the genetic mechanisms responsible for heterosis and the success of maize hybrids. 
Classic overdominance models of heterosis predict that at a single locus two distinct alleles confer heterozygote advantage when combined.
Alternatively, the dominance model predicts that heterosis is driven by dominance effects and the complementation of linked alleles in low-recombination regions (dominance or pseudo-overdominance). 
In the case of true overdominance, selection should thus lead to decreased heterozygosity at the same locus in both populations as complementary haplotypes are fixed in each group \rev{\citep[e.g.][]{guo2014maize}}.
We find \rev{little evidence to support this genetic phenomenon, finding only two 2 cM windows genome-wide in which both populations show significantly reduced heterozygosity.} 
The observed pattern instead favors a dominance model, where fixation of a haplotype in one population simply selects against that same haplotype in the other population. 
\rev{Although strongly deleterious variants were likely purged during the inbreeding process leading to the founder lines, many weakly deleterious alleles can be found segregating at low frequencies among inbreds \citep{Mezmouk2014}.}
Because deleterious alleles will be rare in both populations, most haplotypes in the second population will have a different suite of deleterious variants and will complement the fixed haplotype reasonably well.
\rev{Selection against homozygosity of the fixed haplotype will thus} have little impact on diversity in the second population.

@@

Reciprocal recurrent selection serves as a model for the method of hybrid maize improvement \citep{duvick2004long}.
Similar patterns of diversity and population structure can be seen broadly across North American maize germplasm \citep{van2012historical}, suggesting that genetic drift has most likely played a large role in the current genetic structure of modern maize.  
Nonetheless, breeding has contributed substantially to yield improvements \citep{duvick2005contribution}, and the identification and characterization of loci conferring phenotypic improvement in BSSS and BSCB1 could reveal some of the genetic mechanisms by which maize yield has steadily improved over the decades. 
Several key inbreds in the ‘stiff-stalk’ heterotic group --- B73, B37 and B14 --- were derived from the BSSS population \citep{darrah1986, troyer1999background}. 
%B37 and B14 were derived from cycle 0, and B73 was derived from a half-sib recurrent selection program also started with the BSSS population. 
We examined these three inbreds at the pericentromeric regions listed in Table \ref{tab:fix}, and found that in most cases they carry different haplotypes from those that rose to high frequency in the RRS experiment.  
Near isogenic lines that substitute these centromeric haplotypes from key inbreds could be used to determine if the haplotypes at high frequency in cycle 16 of BSSS provide a phenotypic advantage when crossed with BSCB1. 
Our observation of population-specific selection is consistent with a model of selection for complementation in the Iowa RSS experiment, and the lack of extreme differentiation seen over time across US Corn Belt germplasm may suggest a role for such selection in maize hybrid improvement in general \citep{van2012historical}.

### 2015-08-27 sup1.tex
edits to discussion -- jri finalish draft

\begin{tabular}{ | l | l | l | l | l | l | }

\hline
	\textbf{Simulated 'Hybrid'} & \textbf{Derived Lines Used as Priors} & \textbf{Population} & \textbf{Possible Switches} & \textbf{Switch errors} & \textbf{Rate} \\ \hline
	B11xB67 & all & BSSS & 12195 & 41 & 0.003 \\ \hline
	B17xB44 & all & BSSS & 12230 & 58 & 0.005 \\ \hline
	B39xB37 & all & BSSS & 11485 & 129 & 0.011 \\ \hline
	B43xB69 & all & BSSS & 12266 & 26 & 0.002 \\ \hline
	B73xB72 & all & BSSS & 12043 & 47 & 0.004 \\ \hline
	B78xB94 & all & BSSS & 11658 & 51 & 0.004 \\ \hline
	B89xB84 & all & BSSS & 11517 & 53 & 0.005 \\ \hline
	B11xB67 & cycle 0 & BSSS & 12195 & 45 & 0.004 \\ \hline
	B17xB44 & cycle 0 & BSSS & 12230 & 54 & 0.004 \\ \hline
	B39xB37 & cycle 0 & BSSS & 11485 & 126 & 0.011 \\ \hline
	BB43xB69 & cycle 0 & BSSS & 12266 & 32 & 0.003 \\ \hline
	B42xB54 & all & BSCB1 & 11260 & 38 & 0.003 \\ \hline
	B90xB97 & all & BSCB1 & 6499 & 52 & 0.008 \\ \hline
	B91xB95 & all & BSCB1 & 7830 & 56 & 0.007 \\ \hline
	B99xB97 & all & BSCB1 & 6779 & 52 & 0.008 \\ \hline
	BB42xB54 & cycle 0 & BSCB1 & 11260 & 57 & 0.005 \\\hline
\end{tabular}
	\label{tab:phase_error}  % caption is needed to make this work
\end{table}

@@

\begin{table}
\caption{Evidence of minor contamination between cycles 4 and 8 in BSSS.}
\centering

\subcaption*{Polymorphic SNPs}
\begin{tabular}{ cccc }
	{SNPs polymorphic in cycle} & {But not cycle} & {BSSS} & {BSCB1} \\ \hline
	4 & 0 & 290 & 225 \\ 
	8 & 0 & 2242 & 319 \\
	12 & 0 & 1227 & 181 \\ 
	16 & 0 & 344 & 128 \\ 
	8 & 4 & 4499 & 1081 \\ 
	12 & 4 & 2328 & 669 \\ 
	16 & 4 & 551 & 477 \\ 
	12 & 8 & 1821 & 1405 \\ 
	16 & 8 & 178 & 997 \\ 
	16 & 12 & 542 & 666 \\ 
	0 & Founders & 1202 & 1707 \\ 
	4 & Founders & 822 & 885 \\ 
	8 & Founders & 1201 & 798 \\ 
	12 & Founders & 816 & 550 \\ 
	16 & Founders & 445 & 460 \\ \hline
\end{tabular}
\bigskip
\subcaption*{Heterozygosity of SNPs in cycle 8 but not cycle 4 of the BSSS. }
\begin{tabular}{  lcc }
\multicolumn{1}{l}{} & \multicolumn{1}{l}{BSSS} & \multicolumn{1}{l}{BSCB1}   \\ \cline{2-3}
	Mean & 0.052 & 0.052   \\ 
	Median & 0.028 & 0.028   \\ 
\end{tabular}
		    \label{tab:contamination}  % caption is needed to make this work
\end{table}

>>>>

\begin{tabular}{ cccccc  }
	{Simulated 'Hybrid'} & {Derived Lines Used as Priors} & {Population} & {Possible Switches} & {Switch errors} & {Rate} \\ \hline
	B11xB67 & all & BSSS & 12195 & 41 & 0.003 \\ 
	B17xB44 & all & BSSS & 12230 & 58 & 0.005 \\ 
	B39xB37 & all & BSSS & 11485 & 129 & 0.011 \\ 
	B43xB69 & all & BSSS & 12266 & 26 & 0.002 \\ 
	B73xB72 & all & BSSS & 12043 & 47 & 0.004 \\ 
	B78xB94 & all & BSSS & 11658 & 51 & 0.004 \\ 
	B89xB84 & all & BSSS & 11517 & 53 & 0.005 \\ 
	B11xB67 & cycle 0 & BSSS & 12195 & 45 & 0.004 \\ 
	B17xB44 & cycle 0 & BSSS & 12230 & 54 & 0.004 \\ 
	B39xB37 & cycle 0 & BSSS & 11485 & 126 & 0.011 \\ 
	BB43xB69 & cycle 0 & BSSS & 12266 & 32 & 0.003 \\ 
	B42xB54 & all & BSCB1 & 11260 & 38 & 0.003 \\ 
	B90xB97 & all & BSCB1 & 6499 & 52 & 0.008 \\ 
	B91xB95 & all & BSCB1 & 7830 & 56 & 0.007 \\ 
	B99xB97 & all & BSCB1 & 6779 & 52 & 0.008 \\ 
	BB42xB54 & cycle 0 & BSCB1 & 11260 & 57 & 0.005 \\\hline
\end{tabular}
	\label{tab:phase_error}  % caption is needed to make this work
\end{table}

@@

\begin{table}
\caption{Evidence of minor contamination between cycles 4 and 8 in BSSS.}
\centering
\subcaption*{Polymorphic SNPs}
\begin{tabular}{ cccc }
	{SNPs polymorphic in cycle} & {But not cycle} & {BSSS} & {BSCB1} \\ \hline
	4 & 0 & 290 & 225 \\ 
	8 & 0 & 2242 & 319 \\
	12 & 0 & 1227 & 181 \\ 
	16 & 0 & 344 & 128 \\ 
	8 & 4 & 4499 & 1081 \\ 
	12 & 4 & 2328 & 669 \\ 
	16 & 4 & 551 & 477 \\ 
	12 & 8 & 1821 & 1405 \\ 
	16 & 8 & 178 & 997 \\ 
	16 & 12 & 542 & 666 \\ 
	0 & Founders & 1202 & 1707 \\ 
	4 & Founders & 822 & 885 \\ 
	8 & Founders & 1201 & 798 \\ 
	12 & Founders & 816 & 550 \\ 
	16 & Founders & 445 & 460 \\ \hline
\end{tabular}
\bigskip
\subcaption*{Heterozygosity of SNPs in cycle 8 but not cycle 4 of the BSSS. }
\begin{tabular}{  lcc }
\multicolumn{1}{l}{} & \multicolumn{1}{l}{BSSS} & \multicolumn{1}{l}{BSCB1}   \\ \cline{2-3}
	Mean & 0.052 & 0.052   \\ 
	Median & 0.028 & 0.028   \\ 
\end{tabular}
		    \label{tab:contamination}  % caption is needed to make this work
\end{table}
